---
title: "Redwood Lab Exploration"
author: "Insert Name Here"
format: 
  html:
    code-fold: show
    code-summary: "Show Code"
    code-tools: true
    theme: sandstone
    # theme: yeti
lightbox: true
toc: true
execute: 
  warning: false
  message: false
---

```{r setup, include=FALSE}
# source all functions from R/ folder
for (fname in list.files(here::here("R_solutions"), pattern = "*.R")) {
  source(here::here(file.path("R_solutions", fname)))
}

# path to data
DATA_PATH <- here::here("data")
```

# Load Data

```{r load-data}
# load dates data
dates_orig <- load_dates_data(DATA_PATH)

# load motes data
motes_orig <- load_mote_location_data(DATA_PATH)

# load redwood data
redwood_all_orig <- load_redwood_data(DATA_PATH, source = "all")
redwood_net_orig <- load_redwood_data(DATA_PATH, source = "net")
redwood_log_orig <- load_redwood_data(DATA_PATH, source = "log")
```

## Quick Look

::: {.panel-tabset}

### Dates Data

```{r}
dates_orig
skimr::skim(dates_orig)
```

### Motes Data

```{r}
motes_orig
skimr::skim(motes_orig)
```

### Redwood Data

::: {.panel-tabset .nav-pills}

#### All data

```{r}
redwood_all_orig
skimr::skim(redwood_all_orig)
```

#### Network data

```{r}
redwood_net_orig
skimr::skim(redwood_net_orig)
```

#### Log data

```{r}
redwood_log_orig
skimr::skim(redwood_log_orig)
```

:::

:::

# Intermediary Explorations

Fill in any intermediary explorations or scratch work you'd like to do here (e.g., explorations done to identify issues with the data)

```{r}
plt <- redwood_all_orig |> 
  tidyr::pivot_longer(
    cols = c(humidity, humid_temp, hamatop, hamabot, voltage)
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = value) +
  ggplot2::facet_grid(name ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

# Clean Data

```{r clean-data}
# TODO: fill out cleaning functions in clean.R

# clean dates data
dates_df <- clean_dates_data(dates_orig)

# clean motes data
motes_df <- clean_mote_location_data(motes_orig)

# clean redwood data
redwood_all_df <- clean_redwood_data(redwood_all_orig)
redwood_net_df <- clean_redwood_data(redwood_net_orig)
redwood_log_df <- clean_redwood_data(redwood_log_orig)
```

# Intermediary Explorations

Fill in any intermediary explorations or scratch work you'd like to do here (e.g., explorations to figure out how you can merge the data)

# Merge data

```{r}
# TODO: merge redwood, epoch/dates, and motes data
# Consider writing a "merge_redwood_data()" function (e.g., in your clean.R) to do this merging step
```

# Intermediary Explorations

Fill in any additional explorations you'd like to do after merging the data.
