---
title: "Redwood Lab Exploration"
author: "Insert Name Here"
format: 
  html:
    code-fold: show
    code-summary: "Show Code"
    code-tools: true
    theme: sandstone
    # theme: yeti
    embed-resources: true
lightbox: true
toc: true
execute: 
  warning: false
  message: false
---

```{r setup, include=FALSE}
# source all functions from R/ folder
for (fname in list.files(here::here("R_solutions"), pattern = "*.R")) {
  source(here::here(file.path("R_solutions", fname)))
}

# path to data
DATA_PATH <- here::here("data")
```

# Load Data

```{r load-data}
# load dates data
dates_orig <- load_dates_data(DATA_PATH)

# load motes data
motes_orig <- load_mote_location_data(DATA_PATH)

# load redwood data
redwood_all_orig <- load_redwood_data(DATA_PATH, source = "all")
redwood_net_orig <- load_redwood_data(DATA_PATH, source = "net")
redwood_log_orig <- load_redwood_data(DATA_PATH, source = "log")
```

## Quick Look

::: {.panel-tabset}

### Dates Data

```{r}
dates_orig
skimr::skim(dates_orig)
```

### Motes Data

```{r}
motes_orig
skimr::skim(motes_orig)
```

### Redwood Data

::: {.panel-tabset .nav-pills}

#### All data

```{r}
redwood_all_orig
skimr::skim(redwood_all_orig)
```

#### Network data

```{r}
redwood_net_orig
skimr::skim(redwood_net_orig)
```

#### Log data

```{r}
redwood_log_orig
skimr::skim(redwood_log_orig)
```

:::

:::

# Intermediary Explorations

Fill in any intermediary explorations or scratch work you'd like to do here (e.g., explorations done to identify issues with the data)

```{r}
plt <- redwood_all_orig |> 
  tidyr::pivot_longer(
    cols = c(humidity, humid_temp, hamatop, hamabot, voltage)
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = value) +
  ggplot2::facet_grid(name ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

# Clean Data

```{r clean-data}
# TODO: fill out cleaning functions in clean.R

# clean dates data
dates_df <- clean_dates_data(dates_orig)

# clean motes data
motes_df <- clean_mote_location_data(motes_orig)

# clean redwood data
redwood_all_df <- clean_redwood_data(redwood_all_orig)
redwood_net_df <- clean_redwood_data(redwood_net_orig)
redwood_log_df <- clean_redwood_data(redwood_log_orig)
```

# Intermediary Explorations

Fill in any intermediary explorations or scratch work you'd like to do here (e.g., explorations to figure out how you can merge the data)

# Merge data

```{r}
#| out-width: 100%
redwood_df <- merge_redwood_data(
  dates_data = dates_df,
  motes_data = motes_df,
  redwood_net_data = redwood_net_df,
  redwood_log_data = redwood_log_df
)

redwood_df |> 
  dplyr::slice_head(n = 1000) |> 
  vthemes::pretty_DT()
```

```{r}
skimr::skim(redwood_df)
```

# Intermediary Explorations

Taking a quick look at the various variables over time... 

```{r}
plt <- redwood_df |> 
  tidyr::pivot_longer(
    cols = c(humidity, temp, iPAR, rPAR, voltage)
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = value) +
  ggplot2::facet_grid(name ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

What if we look more closely at temperature and color points by source data?

```{r}
plt <- redwood_df |>
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = temp, color = source) +
  ggplot2::geom_point()
plt
```

Looks like most of the outliers are in the network dataset. We can even toggle between the log/net data and zoom into see what's going on more clostly using `plotly`. (Try double-clicking on the red/blue points in the legend to toggle between the different data sources.)

```{r}
plotly::ggplotly(plt)
```

From this, it looks like the temperature is trailing off to some outrageously large temperature value. What might be going on here? Maybe some network failure? So what if we subset the data to just the network data and color the network temperature values by voltage?

```{r}
plt <- redwood_df |>
  dplyr::filter(source == "net") |> 
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = temp, color = voltage) +
  ggplot2::geom_point()
plt
```

This might suggest that there is some network failure happening that is ultimately leading to erroneous measurements. Temperature doesn't typically fluctuate by that much in a short period of time (recall each epoch is 5 min). Maybe we can use the voltage values in some way to help us do our data cleaning...

Note however that there are still some outliers that don't have a voltage value. We will have to deal with those (and probably other data issues) independently.

Finally, what if we repeated some of these plots for all of the measured variables of interest?

```{r}
plt <- redwood_df |> 
  tidyr::pivot_longer(
    cols = c(humidity, temp, iPAR, rPAR)
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = value, color = source) +
  ggplot2::facet_grid(name ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

Restricted to just the network data and plotting by voltage...

```{r}
plt <- redwood_df |> 
  dplyr::filter(source == "net") |>
  tidyr::pivot_longer(
    cols = c(humidity, temp, iPAR, rPAR)
  ) |> 
  ggplot2::ggplot() +
  ggplot2::aes(x = epoch, y = value, color = voltage) +
  ggplot2::facet_grid(name ~ ., scales = "free_y") +
  ggplot2::geom_point()
plt
```

Again, note that these aren't the only outliers. There are also outliers in the log data, as we've seen above. Further investigation is needed to clean the log data and also to finish cleaning up the network data.

